---
title: Background
output:
  html_document:
    fig_caption: yes
    number_sections: yes
    toc: yes
    df_print: paged
  bookdown::pdf_document2:
    keep_tex: yes
  bookdown::word_document2:
    toc: yes
  bookdown::html_document2: default
bibliography: library.bib
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup methods, include=FALSE}
library(bookdown)
knitr::opts_chunk$set(echo = FALSE, echo = FALSE, message = FALSE, warning = FALSE)
```

```{r}
library(tibble)
Abbreviations <- tibble(
  'Abbreviation' = c('DALY'),
  'Explanation' = c('Disability Adjusted Life Years')
)
```

# Background

Not only are unhealthy diets that lead to malnutrition believed to be
responsible for one in five adult deaths and 15% of the total adult
disability adjusted life years (DALY) globally[@Afshin2019], but our
current food systems, i.e. how we produce, distribute and process food
items, is also taking its toll on the environment. Globally, food
production accounts for 70% of fresh water resources[@Shukla2019](p5),
42-61% of land resources[@Shukla2019](p6) and 21-37% of total greenhouse
gas emissions (GHGE)[@Crippa2021]. Additionally there is growing
awareness of how food production and consumption can negatively impact
biodiversity, pollute air, water and land and increase soil
degradation[@Benton2021; @RodriguezEugenio2018; @Domingo2021;
@Kopittke2019]. These factors have led to the need to develop diets that
are both healthy and environmentally sustainable, so called "Sustainable
Healthy Diets". The World Health Organization (WHO) and the Food and
Agriculture Organization of the United Nations (FAO) defines a
sustainable healthy diet as a diet that: "promote all dimensions of
individuals' health and wellbeing; have low environmental pressure and
impact; are accessible, affordable, safe and equitable; and are
culturally
acceptable."[@FoodandAgricultureOrganizationoftheUnitedNations2019]

While these four aspects of sustainable diets are all important, this
thesis will focus on two of them: Healthiness and environmental impact.

-   Add something about the consequences for children
-   Add something about the sustainability goals?
-   Add something about the pandemic?

## Diet and health

### Consequences of malnutrition

There are three forms of malnutrition that are typically associated with
ill health. Undernutrition (low
weight-for-age/weight-for-height/height-for-age), micronutrient
deficiency or excess, and overnutrition
(overweight/obesity)[@WorldHealthOrganization]. WHO include diet-related
noncommunicable diseases such as cardiovascular disease, type II
diabetes, chronic pulmonary disease and certain types of cancer with
overnutrition. Not only can an unhealthy diet increase the risk of
developing these noncommunicable diseases, it can also worsen prognosis
of those who have already developed them[@Afshin2019; @Wu2016;
@Casas2018; @Steck2020].

Malnutrition is a global health issue, and every country in the world is
affected by at least one form of
malnutrition[@DevelopmentInitiatives2020](p21). Within a population,
household or individual, multiple forms of malnutrition can coexist.
When this happens, it is called the double or triple burden of
malnutrition. For example within a population some can be underweight
while others can be obese and/or suffer from micronutrient deficiencies
or diet-related noncommunicable diseases, and an individual can be obese
due to excess energy intake while also struggling with micronutrient
deficiencies due to low nutrient quality of the foods eaten.

While malnutrition is a global health issue, it affects populations
differently. Undernutrition is more common in low- and
lower-middle-income countries than high-income countries. Globally the
prevalence of undernutrition has decreased, although when looking at
country/regional level there are large variations. On the other hand,
overnutrition is more prevalent in higher-middle-income and high-income
countries, and is increasing globally at an alarming
rate[@DevelopmentInitiatives2020](p51-53).

Micronutrient deficiencies are more common in low- and middle-income
countries than high-income countries[@Bailey2015], but even in high
income countries in Europe and Central Asia micronutrient deficiencies
lead to substantial disease burden and DALYs of 268 per 100 000
population[@Bruins2018]. Regardless of income status, children and women
of childbearing age are at highest risk[@Bailey2015; @Bruins2018]. Of
special concern are iron, iodine, zinc, vitamin A, vitamin D and folate
(vitamin B9), with some regional differences.

-   Add short section on what these nutrients do

-   Vitamin A deficiency is associated with increased mortality and
    morbidity in children under five years of age[@Imdad2017], and is a
    leading cause of preventable childhood blindness[@Wiseman2017]

While it is easy to think that micronutrient deficiencies occur due to a
low intake of food or poor diet quality, there are several factors that
can lead or contribute to micronutrient deficiencies. Such as impaired
digestion or absorption of nutrients, increased excretion or increased
requirements. For example women of childbearing age are at an increased
risk of iron deficiency anemia due to blood loss during menses, and
during pregnancy and lactation women have increased requirements of
several nutrients to support the growing fetus and provide nutrition for
the baby (**SOURCE**).

Malnutrition in any form is especially devastating for children and the
developing fetus. Malnutrition, especially undernutrition, is a
contributing factor in about 45% of death in children below five years
of age[@WorldHealthOrganization], and malnutrition early in life is
associated with impaired cognitive and physical development in addition
to increased risk of developing many diseases in adulthood[@Bailey2015]
(**SOURCES**). There are also concerns that these effects can pass on to
the next generation[@Kaspar2020].

To combat malnutrition at both population and individual level, many
countries and regions, have developed regional/national dietary
guidelines that provide nutritional advice for how to eat to get
adequate energy and nutrients to stay healthy[@Herforth2019]. These
guidelines are created for the general healthy population, and people
with certain diseases may need different and/or additional advice. For
countries without their own national dietary guidelines, WHO has
developed a set of global dietary guidelines[@Herforth2019;
@WorldHealthOrganizationa].

### Healthy dietary patterns

In addition to the national dietary guidelines, other healthy dietary
patterns are described in the literature. Such as the Mediterranean
diet, the Dietary Approaches to Stop Hypertension and the Healthy Eating
Index. These dietary patterns all share common themes: Recommendations
to eat high quality whole foods within ones energy requirements, include
a variety of colorful vegetables (starchy roots and tubers are generally
not counted as a vegetable in these instances) and fruit and other fiber
rich foods such as nuts, legumes and whole grains, consume marine
omega-3 fatty acids and have a low intake of processed foods, added
sugar and salt.

Several studies have found that adherence to the respective populations
dietary guideline is low with the following high to very high
probability of nutrient deficiencies.

-   High/very high probability of Norwegian students to have inadequte
    intake of vitamin A, folate, Vitamin D, iodine, selenium and iron
    [@Valen2020]

-   Page 56-57 in THE STATE OF THE WORLD's BIODIVERSITY THE STATE OF THE
    WORLD's BIODIVERSITTY FOR FOOD AND AGRICULTURE by FAO include
    information about dietary guidelines that take biodiversity into
    account

-   Healthy dietary patterns

-   Country guidelines

-   How healthiness is described to the public -- national guidelines,
    front of pack labels such as Keyhole (Scandinavia), traffic light
    system (UK) and Nutriscore (France, Portugal)

Whole diets are often scored according to a specific dietary pattern,
such as the WHO dietary guidelines, the Mediterranean diet, the New
Nordic Diet, the Healthy Eating Index or others. These scores vary in
their execution, with some providing individual scores for different
parts of the diet, such as vegetable and sodium intake, or a composite
score for all dietary components. The number of different dietary
components vary, and sometimes they can be weighted in relation to each
other. **Provide examples**. While dietary patterns are what is of most
interest when looking for associations between diet and health, it is
also useful to be able to decide if a specific meal or food is a healthy
option. In some cases, dietary guidelines can be used to calculate how
healthy a meal is, such as ... UK CHEF'S VS NORMAL RECIPES SOURCE.
Another option is to use one of the many developed tools to help the
public make healthier food choices. These tools, such as the multiple
traffic light system (later called traffic light system) in the UK,
Nutriscore in France and Keyhole in Scandinavia, provide nutritional
information of a product in an easy to understand manner for the public.
By scoring an individual food or meal on a number of different criteria,
such as amount of saturated fat or added sugar pr. 100 grams or
kilocalorie, important dietary information can be simplified. The
information is presented as a front-of-pack label, that pro These tools,
such as the multiple traffic light system (later called traffic light
system) in the UK, the Nutriscore metric in France and the Keyhole in
Scandinavia,

-IMAGE OF DIFFERENT FOPs-

Throughout the world there are many programs that aim to improve
health/nutrition literacy to help people make good, healthy diet
decisions.

Other ways to help nudge people into more healthy dietary habits is the
use of healthiness indicators on food items bought in stores. These
front-of-pack food labels include information at a glance about a foods
nutritional qualities.

### Healthiness indicators to help the public

Environmentally sustainable diets Climate change is not fair High burden
of malnutrition, also high burden of the effects of climate change

Few national dietary guidelines, that promote healthy dietary patterns,
include information on the environmental impact of the dietary
recommendation7. The Netherlands, France, Germany, Brazil, Qatar, Sweden
and Denmark are exemptions, while the rest of the Nordic countries will
include environmental impact in the new Nordic Nutritional
Recommendations to be published in 2023 (SOURCE). Assessing
environmental impacts of food are usually done by a
life-cycle-assessment ...

Assessing environmental impacts

Reaching consumers Mixed messages in the media regarding food, nutrition
and climate science. People may not know their national food guidelines
(in Norway: "Five a day" is well known, but "three a day" and "three a
week" less so)

## Diet and sustainability

Many of the countries that are the most affected by malnutrition, are
the countries that will experience the worst consequences of climate
change (**SOURCE**).

Globally, deficiency of nutrients that are easily available from animal
based foods, such as iron, vitamin A, zinc and iodine, are
common[@Bailey2015].

### Sustainability indicators to help the public

## Consumer beliefs about diet, health and sustainability

Need for information
