---
title: Background
output:
  html_document:
    fig_caption: yes
    number_sections: yes
    toc: yes
    df_print: paged
  bookdown::pdf_document2:
    keep_tex: yes
  bookdown::word_document2:
    toc: yes
  bookdown::html_document2: default
bibliography: library.bib
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup methods, include=FALSE}
library(bookdown)
knitr::opts_chunk$set(echo = FALSE, echo = FALSE, message = FALSE, warning = FALSE)
```

```{r}
library(tibble)
Abbreviations <- tibble(
  'Abbreviation' = c('DALY'),
  'Explanation' = c('Disability Adjusted Life Years')
)
```

# Background

Not only are unhealthy diets that lead to malnutrition believed to be
responsible for one in five adult deaths and 15% of the total adult
disability adjusted life years (DALY) globally[@Afshin2019], but our
current food systems, i.e. how we produce, distribute and process food
items, is also taking its toll on the environment. Globally, food
production accounts for 70% of fresh water resources[@Shukla2019](p5),
42-61% of land resources[@Shukla2019](p6) and 21-37% of total greenhouse
gas emissions (GHGE)[@Crippa2021]. Additionally there is growing
awareness of how food production and consumption can negatively impact
biodiversity, pollute air, water and land and increase soil
degradation[@Benton2021; @RodriguezEugenio2018; @Domingo2021;
@Kopittke2019]. These factors have led to the need to develop diets that
are both healthy and environmentally sustainable, so called "Sustainable
Healthy Diets". The World Health Organization (WHO) and the Food and
Agriculture Organization of the United Nations (FAO) defines a
sustainable healthy diet as a diet that: "promote all dimensions of
individuals' health and wellbeing; have low environmental pressure and
impact; are accessible, affordable, safe and equitable; and are
culturally
acceptable."[@FoodandAgricultureOrganizationoftheUnitedNations2019]

While these four aspects of sustainable diets are all important, this
thesis will focus on two of them: Healthiness and environmental impact.

-   Add something about the consequences for children
-   Add something about the sustainability goals?
-   Add something about the pandemic?

## Diet and health

### Consequences of malnutrition

There are three forms of malnutrition that are typically associated with
ill health. Undernutrition (low
weight-for-age/weight-for-height/height-for-age), micronutrient
deficiency or excess, and overnutrition
(overweight/obesity)[@WorldHealthOrganization]. WHO include diet-related
noncommunicable diseases such as cardiovascular disease, type II
diabetes, chronic pulmonary disease and certain types of cancer with
overnutrition. Not only can an unhealthy diet increase the risk of
developing these noncommunicable diseases, it can also worsen prognosis
of those who have already developed them[@Afshin2019; @Wu2016;
@Casas2018; @Steck2020].

Malnutrition is a global health issue, and every country in the world is
affected by at least one form of
malnutrition[@DevelopmentInitiatives2020](p21). Within a population,
household or individual, multiple forms of malnutrition can coexist.
When this happens, it is called the double or triple burden of
malnutrition. For example within a population some can be underweight
while others can be obese and/or suffer from micronutrient deficiencies
or diet-related noncommunicable diseases, and an individual can be obese
due to excess energy intake while also struggling with micronutrient
deficiencies due to low nutrient quality of the foods eaten.

While malnutrition is a global health issue, it affects populations
differently. Undernutrition is more common in low- and
lower-middle-income countries than high-income countries. Globally the
prevalence of undernutrition has decreased, although when looking at
country/regional level there are large variations. On the other hand,
overnutrition is more prevalent in higher-middle-income and high-income
countries, and is increasing globally at an alarming
rate[@DevelopmentInitiatives2020](p51-53). There are also differences in
nutrition status between socioeconomic groups within a country, with
undernutrition being more prevalent in low socioeconomic groups than
high socioeconomic group, and the other way around of obesity.

Micronutrient deficiencies are more common in low- and middle-income
countries than high-income countries[@Bailey2015], but even in high
income countries in Europe and Central Asia micronutrient deficiencies
lead to substantial disease burden and DALYs of 268 per 100 000
population[@Bruins2018]. Regardless of income status, children and women
of childbearing age are at highest risk[@Bailey2015; @Bruins2018]. Of
special concern are iron, iodine, zinc, calcium, vitamin A, vitamin D
and folate (vitamin B9), with some regional differences.

All of these micronutrients have several important functions in the
body. Iron is needed to transport oxygen around the body, to synthesize
certain hormones and to maintain healthy cellular functioning and
tissues. Iron is also important for physical growth and neurological
functioning. Dietary iron comes in two forms, heme-iron that is only
found in animal based foods, like red meat and liver, and nonheme-iron
found in both plant- and animal-based foods. Heme-iron is more readily
absorbed in humans, although absorption of nonheme-iron can be improved
in several ways[@NIH-Iron]. A common consequence of iron deficiency is
iron deficiency anemia (IDA), which globally is one of the five leading
causes for years lived with disability[@Vos2017], causing symptoms such
as cognitive impairments and fatigue. It is estimated that 18.8% of
children under five years of age, 20.1% of pregnant women and 16.3% of
nonpregnant women suffer from IDA[@Vos2017].

Iodine is needed for synthesis of thyroid hormones, that regulate
several bodily functions including metabolism. During fetal and infant
development, iodine is important for correct development of the central
nervous system and the skeleton. There are few good food sources of
iodine, with seafood and seaweed being the rare exceptions, leading many
countries to supplementing certain food items with iodine to improve
consumption in the population[@NIH-Iodine]. WHO promote the use of
iodized salt, but also other foods such as milk is used[@Santos2019].
Globally iodine deficiency disorders are a leading cause of preventable
mental retardation, it can also cause goitre, and increases the risk of
perinatal and infant death[@Zimmermann2008].

Zinc is important for the immune system, protein synthesis, wound
healing, cell division, reproductive function, the sense of taste and
smell, and takes part in about 100 enzymatic reactions in the body. Zinc
also plays a part during growth and development, from the fetal stage to
adulthood[@NIH-Zinc]. Zinc can be found in many foods of both animal and
plant origin, but is often more bioavailable from animal based foods. In
plant based sources such as grains and seeds, bioavailability is lowered
due to the presence of phytate, a compound that binds to zinc and
hinders absorption in the human gut. By reducing or removing the phytic
acid from the food, zinc bioavailability improves. This can be done
through processes such as fermenting, soaking and sprouting of grains
and seeds[@Maares2020; @Luo2014]. It is believed zinc deficiency could
be a public health concern in most low- to middle-income
countries[@Gupta2020]. In 2004, zinc deficiency was estimated to be the
cause of 4.4% of deaths in children under five years of age in Latin
America, Africa and Asia, and also accounted for substantial morbidity
from diarrhea, malaria and lower respiratory tract infections in the
same age group and region[@Walker2008].

Calcium is well known for its role in skeletal health and growth, but is
also required for transmission of nerve signals, muscle function and
hormonal secretion[@NIH-Calcium]. Inadequate calcium intake is
associated with poorer bone health, and low bone mineral density which
increase the risk of osteoporosis and fractures. Osteoporosis is a
growing health concern, and it has been estimated that by 2040 274
million people will be at high risk of osteoporotic fractures, a
doubling from 2010[@Oden2015]. More than half of the population at risk
is found in Asia. There are few dietary sources of calcium, with cow's
milk, small fish eaten with bones, and cruciferous vegetables such as
kale and broccoli being among the best sources. However, for calcium to
be absorbed from the diet, the host must have adequate levels of the fat
soluble vitamin D[@NIH-Calcium; @NIH-VitD]. Estimations of average
dietary calcium intake have been found to be low in South, East and
Southeast Asia, and low-moderate in in Africa and South America. In
Europe, and especially Northern Europe average dietary intakes have been
found adequate[@Balk2017]. Many of the countries with suboptimal calcium
intakes also have suboptimal vitamin D status, adding to the problem of
insufficient calcium intake by also hampering absorption of what calcium
is consumed. Suboptimal vitamin D status is also a problem in Northern
European countries with adequate calcium intake[@Wacker2013]. Improving
calcium intake and vitamin D status could possibly improve bone health,
and reduce disease burden of poor bone health[@Balk2017; @Nuti2019].

As already mentioned, vitamin D is essential for proper absorption and
utilization of calcium in the body, and vitamin D deficiency can cause
rickets in children and osteomalacia in adults[@NIH-Calcium; @NIH-VitD].
Other functions of vitamin D include support of the immune system, cell
growth and metabolism, and vitamin D deficiency is associated with a
range of both acute and chronic health conditions. These include
periodontitis, several types of cancer, cardiovascular disease, type II
diabetes, metabolic syndrome, autoimmune disease, asthma and infections
[@Palacios2014; @Roth2018]. There is not established causal
relationships between vitamin D and many of the conditions associated
with vitamin D deficiency, but the potential causal effects of vitamin D
deficiency and potential benefits of increasing vitamin D status are
studied intensively (**SOURCES**. "When not to use meta analysis:
analysing the metaanalyses on vitamin d in critical care", Roth2018).
Vitamin D can be synthesized in the skin by exposure to sunlight, while
some amounts can also be obtained from foods. Supplementation is
necessary in areas with little sun exposure, and for people with
no/little sun exposure due to other reasons, as it is difficult to reach
the required daily intake amounts by diet alone[@Roth2018]. Vitamin D
deficiency is found in both low-, middle- and high-income
countries[@Roth2018].

-   Need some numbers

Vitamin A is a fat soluble vitamin, that like iron comes in multiple
forms. Retinol, or preformed vitamin A, is found in animal based foods
such as dairy, seafood, meat and is found in especially high amounts in
offal such as liver. In plants there are several types of provitamin A,
compounds that can be turned into retinol in the body through enzymatic
reactions. The most important provitamin A is beta-carotene, that gives
many fruits and vegetables their orange color. In the body, vitamin A
supports cell growth, communication between cells, the immune system and
is essential for vision[@NIH-VitA]. Vitamin A deficiency is not common
in high-income countries, but globally it is a leading cause of
preventable childhood blindness[@Wiseman2017], and associated with
increased mortality and morbidity in children under five years of
age[@Imdad2017]. For adults.... **SOURCES**.

Folate is an important water soluble vitamin that is needed for
synthesis of DNA. Inadequate folate intake leads to a reduced rate of
cell division causing a form of anemia called megaloblastic anemia.
Inadequate folate intake during pregnancy can cause neural tube defects
in the developing fetus.

-   Add the deficiency symptoms/diseases! And some epidemiology on the
    consequences of these.

While it is easy to think that micronutrient deficiencies occur due to a
low intake of food or poor diet quality, there are several factors that
can lead or contribute to micronutrient deficiencies. Such as impaired
digestion or absorption of nutrients, increased excretion or increased
requirements. For example women of childbearing age are at an increased
risk of iron deficiency anemia due to blood loss during menses
(**SOURCES**), and during pregnancy and lactation women have increased
requirements of several nutrients to support the growing fetus and
provide nutrition for the baby[@Helsedirektorateta].

Malnutrition in any form is especially devastating for children and the
developing fetus. Malnutrition, especially undernutrition, is a
contributing factor in about 45% of death in children under five years
of age[@WorldHealthOrganization], and malnutrition early in life is
associated with impaired cognitive and physical development in addition
to increased risk of developing many diseases in adulthood[@Bailey2015;
@Peter2014] (**SOURCES**). There are also concerns that these effects
can pass on to the next generation[@Kaspar2020].

To combat malnutrition at both population and individual level, many
countries and regions, have developed regional/national dietary
guidelines that provide nutritional advice for how to eat to get
adequate energy and nutrients to stay healthy[@Herforth2019]. These
guidelines are created for the general healthy population, and people
with certain diseases may need different and/or additional advice. For
countries without their own national dietary guidelines, WHO has
developed a set of global dietary guidelines[@Herforth2019;
@WorldHealthOrganizationa].

### Healthy dietary patterns

In addition to the national dietary guidelines, other healthy dietary
patterns are described in the literature. Such as the Mediterranean
diet, the Dietary Approaches to Stop Hypertension and the Healthy Eating
Index. These dietary patterns all share common themes: Recommendations
to eat high quality whole foods within ones energy requirements, include
a variety of colorful vegetables (starchy roots and tubers are generally
not counted as a vegetable in these instances) and fruit and other fiber
rich foods such as nuts, legumes and whole grains, consume marine
omega-3 fatty acids and have a low intake of processed foods, added
sugar and salt.

Several studies have found that adherence to the respective populations
dietary guideline is low with the following high to very high
probability of nutrient deficiencies.

-   High/very high probability of Norwegian students to have inadequte
    intake of vitamin A, folate, Vitamin D, iodine, selenium and iron
    [@Valen2020]

-   Females and young adults in britain have lower intake than their
    RDIs [@Derbyshire2018]

-   Page 56-57 in THE STATE OF THE WORLD's BIODIVERSITY THE STATE OF THE
    WORLD's BIODIVERSITTY FOR FOOD AND AGRICULTURE by FAO include
    information about dietary guidelines that take biodiversity into
    account

-   Healthy dietary patterns

-   Country guidelines

-   How healthiness is described to the public -- national guidelines,
    front of pack labels such as Keyhole (Scandinavia), traffic light
    system (UK) and Nutriscore (France, Portugal)

Whole diets are often scored according to a specific dietary pattern,
such as the WHO dietary guidelines, the Mediterranean diet, the New
Nordic Diet, the Healthy Eating Index or others. These scores vary in
their execution, with some providing individual scores for different
parts of the diet, such as vegetable and sodium intake, or a composite
score for all dietary components. The number of different dietary
components vary, and sometimes they can be weighted in relation to each
other. **Provide examples**. While dietary patterns are what is of most
interest when looking for associations between diet and health, it is
also useful to be able to decide if a specific meal or food is a healthy
option. In some cases, dietary guidelines can be used to calculate how
healthy a meal is, such as ... UK CHEF'S VS NORMAL RECIPES SOURCE.
Another option is to use one of the many developed tools to help the
public make healthier food choices. These tools, such as the multiple
traffic light system (later called traffic light system) in the UK,
Nutriscore in France and Keyhole in Scandinavia, provide nutritional
information of a product in an easy to understand manner for the public.
By scoring an individual food or meal on a number of different criteria,
such as amount of saturated fat or added sugar pr. 100 grams or
kilocalorie, important dietary information can be simplified. The
information is presented as a front-of-pack label, that pro These tools,
such as the multiple traffic light system (later called traffic light
system) in the UK, the Nutriscore metric in France and the Keyhole in
Scandinavia,

-IMAGE OF DIFFERENT FOPs-

Throughout the world there are many programs that aim to improve
health/nutrition literacy to help people make good, healthy diet
decisions.

Other ways to help nudge people into more healthy dietary habits is the
use of healthiness indicators on food items bought in stores. These
front-of-pack food labels include information at a glance about a foods
nutritional qualities.

### Healthiness indicators to help the public

Environmentally sustainable diets Climate change is not fair High burden
of malnutrition, also high burden of the effects of climate change

Few national dietary guidelines, that promote healthy dietary patterns,
include information on the environmental impact of the dietary
recommendation7. The Netherlands, France, Germany, Brazil, Qatar, Sweden
and Denmark are exemptions, while the rest of the Nordic countries will
include environmental impact in the new Nordic Nutritional
Recommendations to be published in 2023 (SOURCE). Assessing
environmental impacts of food are usually done by a
life-cycle-assessment ...

Assessing environmental impacts

Reaching consumers Mixed messages in the media regarding food, nutrition
and climate science. People may not know their national food guidelines
(in Norway: "Five a day" is well known, but "three a day" and "three a
week" less so)

## Diet and sustainability

Many of the countries that are the most affected by malnutrition, are
the countries that will experience the worst consequences of climate
change (**SOURCE**).

Globally, deficiency of nutrients that are easily available from animal
based foods, such as iron, vitamin A, zinc and iodine, are
common[@Bailey2015].

### Sustainability indicators to help the public

## Consumer beliefs about diet, health and sustainability

Need for information
