# Methods

## Data

To assess the association between different recipes, healthiness of the recipes and their environmental impact, 400 recipes from Norway, and 100 from each of the UK and the US were analysed.

Recipes from Norway and the US came from a database of online recipes collected previously. A randomized selection of these recipes had been chosen for nutritional analyses previously [@Trattner2017] and were reused for this thesis. Recipes came from Klikk.no (n = 100), Tine.no (n = 100), Aperitif.no (n = 100), Kolonial.no (n = 100) (Norway) and Allrecipes.com (n = 100) (US). Since recipe collection Kolonial has been renamed Oda. The UK recipes came from a random selection from an assortment of UK celebrity chef's recipe books previously used to study the healthiness of television chef's recipes [@Howard2012], Baking Made Easy (n = 7), River Cottage Everyday (n = 21), Jamie's Ministry of Food (n = 22), Jamie's 30 Minute Meals (n = 25) and Nigella's Kitchen (n = 25).

## Calculating nutritional content and environmental impact of the recipes

The dataset used included information about each recipe's energy, macronutrient and sodium content. However, for a more detailed macro-and-micronutrient comparison between the countries, and correlation analysis between healthiness, nutrients and sustainability, the ingredients from each recipe were mapped to the 2020 Norwegian Food Composition Database [@NorwegianFoodSafetyAuthority2020], that contains information about the macro- and micronutrient content per 100 g of commonly consumed foods in Norway. If an ingredient in a recipe was not in the Norwegian database, the US equivalent, Agricultural Research Service FoodData Central food database [@foodDataCentral2019], was used. If the ingredient could not be found in either database it was exchanged for a similar ingredient, for example beef liver was used instead of veal liver.

To explore potential beneficial amounts of nutrients in the recipes, the nutrient content was compared to the European Union commissions nutrition claims legislation [@TheEuropeanparliamentandthecounciloftheEuropeanUnion2006]. For micronutrients, a recipe was considered a source of the nutrient if it contained \>15% of the RDI of an adult Norwegian woman [@Helsedirektoratet2014], and a good source if it contained \>30% of the RDI per 100 g. For protein the amounts were 12 and 20 E% respectively, and for dietary fibre 3 and 6 g/MJ.

To calculate a recipe's environmental impact, the SHARP Indicators database was used [@Mertens2019]. The SHARP Indicators database is a public database of certain environmental impacts of different foods, namely greenhouse gas emissions (in kilo CO~2~ equivalents) and land use (in m^2^ pr year) per kg of an ingredient/food. The database includes information on 944 foods commonly consumed in the four European countries Denmark, Czech Republic, Italy and France. The environmental impact in the database comes from life-cycle inventory data on 182 primary products, and many composite foods based on these primary products. These LCAs included the environmental impacts of primary production of the food, and the impact of its packaging, transport to supermarket/consumer, storage at the supermarket/consumer, final preparation and any waste produced throughout.

The nutrient content and environmental impact of each recipe was found by summing the values for all ingredients in the recipe.

Some ingredients were listed in the recipe with either volume units or number of pieces. Before calculating the nutritional content and environmental impact of these ingredients, standardized weight equivalent was found by primarily using the "Weights, measures and portion sizes for foods" database from The Norwegian Directorate of Health [@ØsterholtDalane2015]. If the ingredient was not in this database the U.S. Department of Agriculture, Agricultural Research Service FoodData Central food database [@foodDataCentral2019] was used, the ingredient was found in the online shops of Meny, Kolonial or COOP, or measured at home.

Some recipes included composite ingredients, such as pastry or pizza dough, not present in either nutrient databases or the SHARP Indicator database. For these ingredients a recipe was found through an online search, similar to how the inventors of the SHARP Indicator database looked up composite foods [@Mertens2019]. Nutrient content, CO~2~ emission and land use were calculated for the composite ingredient and normalized to per 100 g of the ingredient, before being added to the recipe like any other ingredient.

For both the sustainability markers and nutritional information, values were normalized to per 100 g of a recipe.

## Recipe healthiness

Recipes were scored on how well they adhered to the dietary guidelines supplied by the WHO [@WHOdietaryguidelines] and the NNR [@NordicCouncilofMinisters2014], see **Table** \@ref(tab:guidelines-points-table). A recipe would receive one point for each recommendation in the guideline it adhered to, giving a total score between 0-6. For example a recipe could score one point for having between 15-30% E% coming from fat. A higher score indicates higher adherence to the guideline.

```{r guidelines-points-table}
#Has problems with %
guidelines_trafficlights %>% select(Feature, `World Health Organization`, `Nordic Nutritional Recommendation`) %>%
  
  #Escape % for latex
  mutate_at(c("World Health Organization", "Nordic Nutritional Recommendation"), ~str_replace(., "E%", "E\\\\%")) %>%
  
  filter(Feature != "Salt") %>%
  rename(Nutrient = Feature) %>%
  kbl(booktabs = TRUE, linesep = "", caption = "Scoring system for the dietary guidelines from World Health Organization and the Nordic countries.", escape = FALSE) %>%
  kable_classic_2(full_width = FALSE) %>%
  kable_styling(latex_options = "hold_position") %>%
  add_header_above(c(' ' = 1, 'Dietary guidelines' = 2))
```

For these analyses, energy providing nutrients fat, saturated fat, carbohydrates, added sugar and protein were recalculated to E%, and dietary fibre to grams per megajoule (MJ), from the original raw values of grams found in the nutrient databases.

Energy percentage of each of the nutrients were found by first calculating the amount of kilocalories provided by each nutrient: multiplying the amount in grams of fat and saturated fat by nine, and for carbohydrates, sugar added sugar and protein by four. This was then divided by the total amount of kilocalories per 100 g and multiplied by 100. Grams of fibre per MJ was found by dividing the amount of fibre in grams by the energy in MJ.

Additionally, the recipes were assessed as to how well they adhered to the UK Food Standard Agency multiple traffic light system (FSA MTL), and the French Nutriscore system (used in several countries in Europe). These scoring systems are front of pack nutrition labels, that have been created to provide consumers at-a-glance information about a single food's nutrient qualities and guide healthier food choices. The FSA MTL looks at a food's fat, saturated fat, sugar and salt content, and gives each category a score between green "low", amber "medium" and red "high" depending on how much of the nutrient the food contains per 100 g, see **Table** \@ref(tab:traffic-lights-points-table). Consumers are advised to choose foods with more green and amber than red [@BritishNutritionFoundation2019]. Foods with mostly green is a "healthier choice", foods with all or mostly amber can be eaten "most of the time", and foods with mostly red should be eaten less often and in low amounts. In a previous study by Howard et al. [@Howard2012] a score of 1 was used for "low", 2 for "medium" and 3 for "high", and a food could score between 4-12, with a lower score being a healthier choice. To be in line with the dietary guideline recommendations where a higher score indicates better adherence, the scoring system by Howard et al. was inverted for this thesis. This means that a total score of seven or more will correspond to a food having mostly amber colored nutrients (three amber and one red) and can be eaten "most of the time".

```{r traffic-lights-points-table}
guidelines_trafficlights %>%
      select(Feature, `Green/low`, `Amber/medium`, `Red/high`) %>%
      filter(!Feature %in% c('Carbohydrate', 'Dietary fibre', 'Protein')) %>%
  #LaTeX less than or equal to sign
  mutate(across(everything(), ~str_replace(., "<=", "$\\\\leq$"))) %>%
  rename(Nutrient = Feature,
         `Healthier choice` = `Green/low`,
         `Can be eaten most of the time` = `Amber/medium`,
         `Eat less often` = `Red/high`) %>%
  
  kbl(booktabs = TRUE, linesep = "", escape = FALSE, caption = "Scoring system for the UK Food Standard Agency multiple traffic light system") %>%
  add_header_above(c(" ", "Green/Low" = 1, "Amber/Medium" = 1, "Red/High" = 1)) %>%
  kable_styling(latex_options = "hold_position") %>%
  kable_classic_2()
```

The Nutriscore is similar, but it builds on the FSA MTL by also taking into account the amount of energy and protein in a food, in addition to the amount of fibre, fruit, vegetables, pulses and nuts and the healthy fats olive oil, rapeseed oil and walnut oil [@Chantal2017; @Hercberg2021]. In the Nutriscore system, a food can receive between 0-10 points for each of the "disqualifying" categories energy, added sugar, saturated fat and sodium, and 0-5 points for the "qualifying" categories of protein, fibre, and the percentage of fruit, vegetables, pulses and nuts and the healthy fats olive oil, rapeseed oil and walnut oil per 100 g of the food [@NutriscoreGuidelines]. In contrast from the FSA MTL, the Nutriscore is a summary score and does not provide the consumer with the individual scores from each category.

The points from the qualifying categories are subtracted from the points from the disqualifying categories, see **Table** \@ref(tab:nutriscore-points-table), giving a total score between -15 to +40, and a lower score indicates a more healthy and nutritious food. The points correspond to a colored grading system between a dark green A (most healthy) and a dark orange E (least healthy). Consumption of foods that score green/A, is encouraged, while consumption of foods that score dark orange/E is advised to be limited [@Dreano-Trecant2020]. As with the FSA MTL, an inverted scoring system is used in the analyses in this thesis. An inverted score of minus two or more, corresponding to a light green B or dark green A, was chosen as the cut-off for a food that could be eaten often as part of a healthy diet. This score has also previously been used to study if the Nutriscore is in alignment with national FBDGs [@Ahmed2020].

```{r, eval = knitr::is_latex_output(), results='asis'}
\begin{landscape}
```
```{r nutriscore-points-table}
if(knitr::is_latex_output()){
nutriscore_points %>%
  mutate(across(everything(), ~str_replace(., "<=", "$\\\\leq$"))) %>%
  #Add units
  mutate(`Energy density (kJ/100g)` = paste0(`Energy density (kJ/100g)`, " kJ"),
         
         across(c(`Added sugars (g/100g)`, `Saturated fat (g/100g)`),
                ~paste0(., " g")),
         
         across(c(`Fibre (g/100g)`, `Protein (g/100g)`), ~case_when(!str_detect(., "-") ~ paste0(., " g"),
                                                                   TRUE ~ .)),
         `Sodium (mg/100g)` = paste0(`Sodium (mg/100g)`, " mg"),
         
         `Fruits, vegetables, pulses, nuts and rapeseed, walnut and olive oils (%)` = case_when(
           !str_detect(`Fruits, vegetables, pulses, nuts and rapeseed, walnut and olive oils (%)`, "-") ~ paste0(`Fruits, vegetables, pulses, nuts and rapeseed, walnut and olive oils (%)`, " \\%"),
           TRUE ~ `Fruits, vegetables, pulses, nuts and rapeseed, walnut and olive oils (%)`)
         ) %>%
  
    #KableExtra
kbl(booktabs = TRUE, linesep = "", format = "latex", escape = FALSE, caption = "Scoring system for Nutriscore.",
      col.names = c("Points", "Energy density", "Added sugars", "Saturated fat", "Sodium", "Fruits, vegetables, pulses, nuts and rapeseed, walnut and olive oils", "Dietary fibre", "Protein")
      ) %>%
  #kable_classic_2(full_width = FALSE) %>%
  kable_styling(latex_options = c("scale_down", "hold_position")) %>%
  add_header_above(c(' ' = 1, 'Disqualifying' = 4, 'Qualifying' = 3)) %>%
  #Linebreak for long column names
  column_spec(2:4, width = "3cm") %>%
  column_spec(c(5, 7, 8), width = "2.5cm") %>%
  column_spec(6, width = "4cm")
  
} else if(knitr::is_html_output()) {
  
  nutriscore_points %>%
  #Add units
  mutate(`Energy density (kJ/100g)` = paste0(`Energy density (kJ/100g)`, " kJ"),
         
         across(c(`Added sugars (g/100g)`, `Saturated fat (g/100g)`),
                ~paste0(., " g")),
         
         across(c(`Fibre (g/100g)`, `Protein (g/100g)`), ~case_when(!str_detect(., "-") ~ paste0(., " g"),
                                                                   TRUE ~ .)),
         `Sodium (mg/100g)` = paste0(`Sodium (mg/100g)`, " mg")
         ) %>%
  
    #KableExtra
kbl(booktabs = TRUE, linesep = "", escape = FALSE, caption = "Scoring system for Nutriscore.",
      col.names = c("Points", "Energy density", "Added sugars", "Saturated fat", "Sodium", "Fruits, vegetables, pulses, nuts and rapeseed, walnut and olive oils", "Dietary fibre", "Protein")
      ) %>%
  kable_classic_2(full_width = FALSE) %>%
  kable_styling(latex_options = c("scale_down", "hold_position")) %>%
  add_header_above(c(' ' = 1, 'Disqualifying' = 4, 'Qualifying' = 3)) %>%
  #Linebreak for long column names
  column_spec(2:4, width = "3cm") %>%
  column_spec(c(5, 7, 8), width = "2.5cm") %>%
  column_spec(6, width = "4cm")
}

```
```{r, eval = knitr::is_latex_output(), results='asis'}
\end{landscape}
```

The percentage of fruit, vegetables, pulses, nuts and healthy fats were found by summing the amounts of ingredients found in the food groups "Fruit and fruit products", "Vegetable and vegetable products", "Fruit/vegetable juice and nectar" and "Legumes, nuts, seeds" from the SHARP Indicator database, in addition to the amounts of olive, rapeseed and walnut oil per 100 g of the recipe.

## Statistical analyses

### Missing data

Ingredients with no amounts were replaced with the mean value of that ingredient in all other recipes, normalized to per 100 g of the recipe. If this was not possible, ingredients were left out of the analyses. Ingredients that could not be found in the nutrient or sustainability databases were if possible exchanged for a similar ingredient, or left out. If more than 10% of the recipe in weight could not be mapped to either the food composition table or SHARP Indicators database, the recipe was left out.

### Significant differences between countries

Significant differences in energy per 100 g, macronutrient content in percentage of energy, micronutrient content in percentage of the RDI of an adult woman in Norway [@Helsedirektoratet2014], healthiness scores and sustainability indicators between recipes from the different countries were assessed using the Kruskal-Wallis test. The Kruskal-Wallis test is a non-parametric test to look for differences between more than two groups [@Ostertagova2014]. A significant result means that there is a significant difference between at least two groups, but the test cannot say which groups. To identify the groups that were different from each other a *post-hoc* pair-wise comparison was done using the Dunn's test [@Dinno2015]. To account for multiple testing the Benjamini Hochberg (BH) method was used [@Benjamini1995]. An adjusted *p*-value \< 0.05 was considered significant.

All tests were done using the rstatix package v. 0.7.0 [@Kassambara2021].

### Correlation analysis

To explore possible correlation between variables, the nonparametric test Spearman's rho [@SpearmanCorr] was calculated using the Psych library v. 2.1.6 [@Raiche2013] corr.test function. This was done for the recipes from each country individually and for the pooled data from all countries.

P values were calculated and corrected for multiple testing using the BH method and an adjusted *p*-value \< 0.05 was considered significant.
